---
title: 'BC Chronic Disease: Temporal Modeling EDA'
date: "May 17, 2022"
author: 'Jennifer Hoang'
output:
  html_document:
    df_print: paged
---

```{r Load Libraries, include=FALSE}
library(tidyverse)
library(ggplot2)
library(here)
```

```{r Load Data, include=FALSE}
path <- here("data", "processed")

inc_rate_df <- read_csv(here(path, "incidence_rate_combined.csv"), col_types = cols(CLNT_GENDER_LABEL = col_factor()))
```

## Overview of Disease Incidence Trends with Local Polynomial Regression (LOESS)

To get an overview of the trends for different diseases, 
I'll look at the mean standardized incidence rate among all the available CHSA first. 
I'll use LOESS as a smoothing model with the default span parameter.

```{r Calculate Mean Incidence per Disease, echo=FALSE, message=FALSE}
mean_inc <- inc_rate_df |>
  group_by(DISEASE, YEAR) |>
  summarise(
    mean_crude_rate = mean(CRUDE_RATE_PER_1000, na.rm = TRUE),
    mean_std_rate = mean(STD_RATE_PER_1000, na.rm = TRUE)
  )

diseases <- unique(mean_inc$DISEASE)
```

```{r Plot per Disease, echo=FALSE, fig.show="hold", out.width="25%"}
disease_plots = list()

for (d in diseases) {
  disease_data <- mean_inc |> 
    filter(DISEASE == d) 
  
  disease_plots[[d]] = ggplot(disease_data, aes(x = YEAR, y = mean_std_rate, group = 1)) +
    geom_point() +
    geom_line() +
    labs(x = "Year", y = "Mean Std Incidence Rate", title = toString(d)) +
    geom_smooth(formula = y ~ x, method = "loess", fill = "lightblue")
  
  print(disease_plots[[d]])
}
```

We can see that each disease has unique trends that are often non-linear. 
Local regression (loess) does a good job of identifying local trends and could be a good baseline model.

Taking a closer look, however, loess occasionally produces confidence intervals 
with values less 0 that do not make sense, such as a negative incidence rate. 
For MS, the disease incidence rate is quite low, and this sometimes produces negative CIs.

### Region-Specific and Disease-Specific Trends

Let's compare the confidence intervals of the loess model to the 95% CIs calculated in the dataset. 
We'll check two different CHSA with a smaller and a larger population.

```{r Plot Trend in Specific Disease and Region, echo=FALSE, warning=FALSE, fig.show="hold", out.width="50%"}
plot_trend <- function(data, disease, chsa, span = 0.75) {
  plot_data <- inc_rate_df |> 
    filter(DISEASE == disease & HEALTH_BOUNDARIES == chsa)
  
  ggplot(plot_data, aes(x = YEAR, y = STD_RATE_PER_1000, group = 1)) +
    geom_point() +
    geom_line() +
    geom_ribbon(aes(ymin = STD_LCL_95, ymax = STD_UCL_95), alpha = 0.5) +
    labs(x = "Year", y = "Standardized Incidence Rate",
         title = paste(disease, "in", chsa)) +
    geom_smooth(formula = y ~ x, method = "loess", fill = "lightblue", span = span)
  
}

plot_trend(inc_rate_df, "MS", "1110 Fernie")
plot_trend(inc_rate_df, "MS", "2223 Burnaby Southwest")
plot_trend(inc_rate_df, "GOUT", "1110 Fernie")
plot_trend(inc_rate_df, "GOUT", "2223 Burnaby Southwest")

```

The confidence interval bands tend to be more narrow. 
The loess model uses t-based confidence interval approximation based on the `geom_smooth` documentation.

### Incidence Rates in one CHSA

Here are the incidence rates for all diseases in 1110 Fernie. 
Again, loess seems to work well except for rare diseases.

```{r Region-Specific Plots, echo=FALSE, fig.show="hold", out.width="25%"}
for (d in diseases) {
  p <- plot_trend(inc_rate_df, d, "1110 Fernie")
  print(p)
}
```
