---
title: 'BC Chronic Disease: Temporal Modeling with INLA'
date: "May 17, 2022"
author: 'Jennifer Hoang'
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(ggplot2)
library(INLA)
library(here)
```

```{r Load Data, include=FALSE}
path <- here("General", "Data")

inc_rate_df <- read_csv(here(path, "incidence_rate_combined.csv"), col_types = cols(CLNT_GENDER_LABEL = col_factor()))
hsc_prev_df <- read_csv(here(path, "hsc_prevalence_combined.csv"), col_types = cols(CLNT_GENDER_LABEL = col_factor()))
life_prev_df <- read_csv(here(path, "life_prevalence_combined.csv"), col_types = cols(CLNT_GENDER_LABEL = col_factor()))

x_year <- unique(inc_rate_df$YEAR)
```

## Comparing Random Walk Models

We will try a Gamma prior distribution for the rates, which are continuous.
One challenge is that Gamma does not support 0 values, so rare diseases can't be modeled.

```{r Functions to Fit and Plot Models}
#' Fit a random walk 1 model with STD_RATE_PER_1000 as response variable and 
#' YEAR as explanatory variable with INLA. Assumes Gamma likelihood.
#'
#' @param data data frame containing original data
#' @return data frame with model results
fit_inla_rw1 <- function(data) {
  model <- inla(STD_RATE_PER_1000 ~ 1 + f(YEAR, model = "rw1"),
    data = data,
    family = "gamma",
    control.compute = list(dic = TRUE, waic = TRUE)
  )

  tibble(
    x = x_year,
    y_fitted = model$summary.fitted$mean,
    ci_95_ll = model$summary.fitted$`0.025quant`,
    ci_95_ul = model$summary.fitted$`0.975quant`,
    dic = model$dic$dic,
    waic = model$waic$waic,
    model = "rw1"
  )
}

#' Fit a random walk 2 model with STD_RATE_PER_1000 as response variable and 
#' YEAR as explanatory variable with INLA. Assumes Gamma likelihood.
#'
#' @param data data frame containing original data
#' @return data frame with model results
fit_inla_rw2 <- function(data) {
  model <- inla(STD_RATE_PER_1000 ~ 1 + f(YEAR, model = "rw2"),
    data = data,
    family = "gamma",
    control.compute = list(dic = TRUE, waic = TRUE)
  )

  tibble(
    x = x_year,
    y_fitted = model$summary.fitted$mean,
    ci_95_ll = model$summary.fitted$`0.025quant`,
    ci_95_ul = model$summary.fitted$`0.975quant`,
    dic = model$dic$dic,
    waic = model$waic$waic,
    model = "rw2"
  )
}

#' Fit a loess model with STD_RATE_PER_1000 as response variable and 
#' YEAR as explanatory variable. Confidence intervals generated
#' using t-distribution approximation
#'
#' @param data data frame containing original data
#' @return data frame with model results
fit_loess <- function(data) {
  model <- loess(STD_RATE_PER_1000 ~ YEAR, data = data)
  model_predict <- predict(model, data, se = TRUE)

  t <- qt(p = 0.975, df = model_predict$df)

  tibble(
    x = x_year,
    y_fitted = model_predict$fit,
    ci_95_ll = model_predict$fit - t * model_predict$se.fit,
    ci_95_ul = model_predict$fit + t * model_predict$se.fit,
    dic = NA,
    waic = NA,
    model = "loess"
  )
}

#' Generate line plot to compare models and confidence intervals
#'
#' @param data data frame containing original data
#' @param model_df data frame containing model results
#' @param disease character corresponding to entry in DISEASE column
#' @param chsa character corresponding to entry in HEALTH_BOUNDARIES column
#' @param rate_label character (label for Y axis)
#' @return ggplot object
plot_models <- function(data, model_df, disease, chsa, rate_label) {
  cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
  
  ggplot(aes(x = YEAR, y = STD_RATE_PER_1000), data = data) +
    geom_point() +
    geom_ribbon(aes(x = x, y = y_fitted, ymin = ci_95_ll, ymax = ci_95_ul, fill = model), alpha = 0.2, data = model_df) +
    geom_line(aes(x = x, y = y_fitted, color = model), data = model_df, linetype = 2) +
    labs(
      x = "Year", y = rate_label,
      title = paste(disease, "in", chsa)
    ) +
    scale_fill_manual(values=cbPalette) +
    scale_colour_manual(values=cbPalette)
}
```


```{r Compare 3 Models for 1 Outcome, fig.show="hold", out.width="25%"}
#' Generate line plot to compare loess, random walk 1, and random walk 2 models
#' Only fits data without 0 values due to limitations with Gamma likelihood.
#'
#' @param data data frame
#' @param chsa character corresponding to entry in HEALTH_BOUNDARIES column
#' @param rate_label character (label for Y axis)
#' @return ggplot object
compare_models <- function(data, chsa, rate_label) {
  diseases <- unique(data$DISEASE)
  
  for (d in diseases) {
    disease_data <- data |>
      filter(DISEASE == d & HEALTH_BOUNDARIES == chsa)

    if (all(disease_data$STD_RATE_PER_1000 > 0)) {
      model_df <- rbind(fit_inla_rw1(disease_data), fit_inla_rw2(disease_data), fit_loess(disease_data))
      
      p <- plot_models(disease_data, model_df, d, chsa, rate_label)
      print(p)
    }
  }
}
```

### Age-Standardized Life Prevalence in Fernie

Note: not all plots are generated for all diseases because some rare diseases have 0 entries.
```{r}
compare_models(inc_rate_df, "1110 Fernie", "Age-Std Incidence Rate")
```

### Age-Standardized Life Prevalence in Fernie
```{r}
compare_models(life_prev_df, "1110 Fernie", "Age-Std Life Prevalence Rate")
```

### Standardized HSC Prevalence in Fernie
```{r}
compare_models(hsc_prev_df, "1110 Fernie", "Age-Std HSC Prevalence Rate")
```




