---
title: 'BC Chronic Disease: Temporal Modeling with INLA (Tweedie)'
date: "June 1, 2022"
author: 'Jennifer Hoang'
output:
  html_document:
    df_print: paged
    code_folding: hide
---

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
library(INLA)
library(here)
```

```{r Load Data, message=FALSE}
path <- here("data", "processed")

inc_rate_df <- read_csv(here(path, "incidence_rate_combined.csv"), col_types = cols(CLNT_GENDER_LABEL = col_factor()))
hsc_prev_df <- read_csv(here(path, "hsc_prevalence_combined.csv"), col_types = cols(CLNT_GENDER_LABEL = col_factor()))
life_prev_df <- read_csv(here(path, "life_prevalence_combined.csv"), col_types = cols(CLNT_GENDER_LABEL = col_factor()))

x_year <- unique(inc_rate_df$YEAR)
```

## Tweedie Likelihood for Semi-continuous Data

We will try a Tweedie likelihood for the response variable to accommodate 0 values in the continuous response variable. 
The Tweedie distribution family has 3 parameters, including the mean, variance, and index parameter (p), 
and it has been used in other healthcare contexts, such as utilization costs [Kurz 2017](https://doi.org/10.1186/s12874-017-0445-y).

When 1 \< p \< 2, the distribution is a compound Poisson-gamma mixture model, which is continuous with a non-negative mass at 0 (<https://www.rdocumentation.org/packages/statmod/versions/1.4.33/topics/tweedie>). 
When p = 1, the distribution is equivalent to a Poisson distribution. 
When p = 2, the distribution is equivalent to a gamma distribution.

However, the Tweedie distribution implementation in INLA is currently experimental and subject to changes.
Due to the experimental nature and some issues with fitting the model on the data below,
the models generated with this distribution may require more careful review.

```{r Functions to Fit and Plot Models}
#' Fit a random walk 1 model with STD_RATE_PER_1000 as response variable and 
#' YEAR as explanatory variable with INLA. Assumes Tweedie likelihood.
#'
#' @param data data frame containing original data
#' @return data frame with model results
fit_inla_rw1 <- function(data) {
  model <- inla(STD_RATE_PER_1000 ~ 1 + f(YEAR, model = "rw1"),
    data = data,
    family = "tweedie",
    control.compute = list(dic = TRUE, waic = TRUE)
  )

  model_results <- tibble(
    DISEASE = data$DISEASE,
    HEALTH_BOUNDARIES = data$HEALTH_BOUNDARIES, 
    x = x_year,
    y_fitted = model$summary.fitted$mean,
    ci_95_ll = model$summary.fitted$`0.025quant`,
    ci_95_ul = model$summary.fitted$`0.975quant`,
    dic = model$dic$dic,
    waic = model$waic$waic,
    model = "rw1"
  )
  
  model_results
}

#' Fit a random walk 2 model with STD_RATE_PER_1000 as response variable and 
#' YEAR as explanatory variable with INLA. Assumes Tweedie likelihood.
#'
#' @param data data frame containing original data
#' @return data frame with model results
fit_inla_rw2 <- function(data) {
  model <- inla(STD_RATE_PER_1000 ~ 1 + f(YEAR, model = "rw2"),
    data = data,
    family = "tweedie",
    control.compute = list(dic = TRUE, waic = TRUE)
  )

  model_results <- tibble(
    DISEASE = data$DISEASE,
    HEALTH_BOUNDARIES = data$HEALTH_BOUNDARIES, 
    x = x_year,
    y_fitted = model$summary.fitted$mean,
    ci_95_ll = model$summary.fitted$`0.025quant`,
    ci_95_ul = model$summary.fitted$`0.975quant`,
    dic = model$dic$dic,
    waic = model$waic$waic,
    model = "rw2"
  )
  
  model_results
}

#' Fit a loess model with STD_RATE_PER_1000 as response variable and 
#' YEAR as explanatory variable. Confidence intervals generated
#' using t-distribution approximation
#'
#' @param data data frame containing original data
#' @return data frame with model results
fit_loess <- function(data) {
  model <- loess(STD_RATE_PER_1000 ~ YEAR, data = data)
  model_predict <- predict(model, data, se = TRUE)

  t <- qt(p = 0.975, df = model_predict$df)

  model_results <- tibble(
    DISEASE = data$DISEASE[1],
    HEALTH_BOUNDARIES = data$HEALTH_BOUNDARIES[1],
    x = x_year,
    y_fitted = model_predict$fit,
    ci_95_ll = model_predict$fit - t * model_predict$se.fit,
    ci_95_ul = model_predict$fit + t * model_predict$se.fit,    
    dic = NA,
    waic = NA,
    model = "loess",
  )
  
  model_results
}

#' Generate line plot to compare models and confidence intervals
#'
#' @param data data frame containing original data
#' @param model_df data frame containing model results
#' @param disease character corresponding to entry in DISEASE column
#' @param chsa character corresponding to entry in HEALTH_BOUNDARIES column
#' @param rate_label character (label for Y axis)
#' @return ggplot object
plot_models <- function(data, model_df, disease, chsa, rate_label) {
  cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
  
  ggplot(aes(x = YEAR, y = STD_RATE_PER_1000), data = data) +
    geom_point() +
    geom_ribbon(
      aes(x = x, y = y_fitted, ymin = ci_95_ll, ymax = ci_95_ul, fill = model), 
      alpha = 0.2, data = model_df
    ) +
    geom_line(aes(x = x, y = y_fitted, color = model), data = model_df, linetype = 2) +
    labs(
      x = "Year", y = rate_label,
      title = paste(disease, "in", chsa)
    ) +
    scale_fill_manual(values=cbPalette) +
    scale_colour_manual(values=cbPalette)
}
```

```{r Compare 3 Models for 1 Outcome, fig.show="hold", out.width="25%"}
#' Generate line plot to compare loess, random walk 1, and random walk 2 models
#' Only fits data without 0 values due to limitations with Gamma likelihood.
#'
#' @param data data frame
#' @param chsa character corresponding to entry in HEALTH_BOUNDARIES column
#' @param rate_label character (label for Y axis)
#' @return ggplot object
compare_models <- function(data, chsa, rate_label) {
  diseases <- unique(data$DISEASE)
  all_model_df <- tibble()
  plot_df <- tibble()

  for (d in diseases) {
    disease_data <- data |>
      filter(DISEASE == d & HEALTH_BOUNDARIES == chsa)

    if (any(disease_data$STD_RATE_PER_1000 == 0)) {
      tryCatch(
        {plot_df <- rbind(fit_inla_rw1(disease_data), fit_inla_rw2(disease_data), fit_loess(disease_data))
        
        p <- plot_models(disease_data, plot_df, d, chsa, rate_label)
        print(p)
        },
        error = function(e) {
          print(paste("Error with", d, chsa))
        }
      )
    }
    all_model_df <- rbind(all_model_df, plot_df)
  }
  all_model_df
}
```

### Standardized HSC Prevalence in Fernie

```{r fig.show="hold", out.width="50%"}
hsc_fernie <- compare_models(hsc_prev_df, "1110 Fernie", "Age-Std HSC Prevalence Rate")
```

```{r Wrangle Model Criteria}
wrangle_model_results <- function(data) {
  data |>
  select(DISEASE, HEALTH_BOUNDARIES, model, dic, waic) |>
  filter(model != "loess") |>
  unique() |>
  pivot_wider(names_from = model,
              values_from = c(dic, waic)) |>
  mutate(best_model_dic = ifelse(dic_rw1 < dic_rw2, "rw1", "rw2"),
         best_model_waic = ifelse(waic_rw1 < waic_rw2, "rw1", "rw2")) |>
  mutate_if(is.numeric, round, 2) 
}

wrangle_model_results(hsc_fernie)
```

### Age-Standardized Incidence Rate in Fernie

```{r fig.show="hold", out.width="50%"}
inc_fernie <- compare_models(inc_rate_df, "1110 Fernie", "Age-Std Incidence Rate")

wrangle_model_results(inc_fernie)
```

### Age-Standardized Incidence Rate in Vanderhoof Rural

There are many zero values in this CHSA, so we will investigate how the model performs here.

```{r fig.show="hold", out.width="50%"}
inc_vanderhoof <- compare_models(inc_rate_df, "5232 Vanderhoof Rural", "Age-Std Incidence Rate")

wrangle_model_results(inc_vanderhoof)
```
