---
title: "BC Chronic Disease: Temporal Model Plots"
output: html_document
date: '2022-05-31'
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r Load Libraries, message=FALSE}
library(shiny)
library(tidyverse)
library(ggplot2)
library(here)
```


```{r Load Data and Functions, message=FALSE}
#' Read and concatenate all csv files in a folder
#'
#' @param path to folder containing csv files
#' @return dataframe with concatenated data
load_files <- function(path) {
  combined_df <- data.frame()

  for (file in list.files(here(path))) {
    new_df <- read_csv(here(path, file))
    combined_df <- rbind(combined_df, new_df)
  }
  combined_df
}

#' Wrangle original dataset to add numeric year column
#'
#' @param dataframe with original data
#' @return dataframe with year column
wrangle_orig_df <- function(data) {
  data |>
    mutate(YEAR = as.numeric(substr(FISC_YR_LABEL, 4, 7)))
}

#' Create line plot with smoothing model with original point
#' observations and confidence intervals
#'
#' @param model_data dataframe with fitted model results
#' @param orig_data dataframe with original data
#' @param title y_title title for y-axis (character)
#' @param input_disease name of disease for plot title (character)
#' @param input_region name of region for plot title (character)
#' @return ggplot object
line_plot <- function(model_data, orig_data, y_title, input_disease, input_region) {
  model_data |>
    ggplot(aes(x = year, y = y_fitted)) +
    geom_ribbon(
      data = orig_data,
      aes(x = YEAR, y = STD_RATE_PER_1000, ymin = STD_LCL_95, ymax = STD_UCL_95),
      alpha = 0.2,
      fill = "lightblue"
    ) +
    geom_point(data = orig_data, aes(x = YEAR, y = STD_RATE_PER_1000)) +
    geom_line() +
    geom_ribbon(
      aes(ymin = ci_95_ll, ymax = ci_95_ul),
      alpha = 0.2
    ) +
    geom_line(aes(x = year, y = y_fitted), linetype = 2) +
    labs(
      x = "Fiscal Year", y = y_title,
      title = paste(input_disease, "in", input_region)
    ) +
    theme_grey(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    scale_x_continuous(breaks = year_labels, labels = fiscal_year_labels) +
    expand_limits(y = 0)
}

# Load model data
hsc_prev_df <- load_files(here("results", "model", "HSCPrevalence"))
inc_rate_df <- load_files(here("results", "model", "IncidenceRate"))
life_prev_df <- load_files(here("results", "model", "LifePrevalence"))

# Load original data set for 95% confidence interval comparison
orig_hsc_prev_df <- load_files(here("data", "Data_T_CHSA", "HSCPrevalence")) |>
  wrangle_orig_df()
orig_inc_rate_df <- load_files(here("data", "Data_T_CHSA", "IncidenceRate")) |>
  wrangle_orig_df()
orig_life_prev_df <- load_files(here("data", "Data_T_CHSA", "LifePrevalence")) |>
  wrangle_orig_df()


year_labels <- unique(orig_hsc_prev_df$YEAR)
fiscal_year_labels <- unique(orig_hsc_prev_df$FISC_YR_LABEL) |>
  substr(4, 12)
```

## Age-Standardized HSC Prevalence Rate

```{r HSC Prev Rate Plot}
# Input -----------------------------------------------------------------------
inputPanel(
  selectInput("hsc_region",
    label = "CHSA",
    choices = sort(unique(orig_hsc_prev_df$HEALTH_BOUNDARIES)),
    selected = "1110 Fernie"
  ),
  selectInput("hsc_disease",
    label = "Disease",
    choices = unique(orig_hsc_prev_df$DISEASE),
    selected = "AMI_EPI"
  )
)

# Reactive Elements -----------------------------------------------------------
hsc_selected_data <- reactive({
  hsc_prev_df |>
    dplyr::filter(HEALTH_BOUNDARIES == input$hsc_region & DISEASE == input$hsc_disease)
})

orig_hsc_selected_data <- reactive({
  orig_hsc_prev_df |>
    dplyr::filter(HEALTH_BOUNDARIES == input$hsc_region & DISEASE == input$hsc_disease)
})

# Output -----------------------------------------------------------------------
renderPlot({
  line_plot(hsc_selected_data(), orig_hsc_selected_data(),
    y_title = "Age Standardized HSC Prevalence Rate (per 1000)",
    input$hsc_disease, input$hsc_region
  )
})

renderText({
  paste("The selected model was:", hsc_selected_data()$model[1])
})
```
Ribbons: Blue = 95% confidence interval, 
Grey = 95% credible interval if RW1 or RW2 selected, 
otherwise 95% confidence interval if loess selected

## Age-Standardized Incidence Rate

```{r Inc Rate Plot}
# Input -----------------------------------------------------------------------
inputPanel(
  selectInput("inc_region",
    label = "CHSA",
    choices = sort(unique(orig_inc_rate_df$HEALTH_BOUNDARIES)),
    selected = "1110 Fernie"
  ),
  selectInput("inc_disease",
    label = "Disease",
    choices = unique(orig_inc_rate_df$DISEASE),
    selected = "AMI"
  )
)

# Reactive Elements -----------------------------------------------------------
inc_selected_data <- reactive({
  inc_rate_df |>
    dplyr::filter(HEALTH_BOUNDARIES == input$inc_region & DISEASE == input$inc_disease)
})

orig_inc_selected_data <- reactive({
  orig_inc_rate_df |>
    dplyr::filter(HEALTH_BOUNDARIES == input$inc_region & DISEASE == input$inc_disease)
})

# Output -----------------------------------------------------------------------
renderPlot({
  line_plot(inc_selected_data(), orig_inc_selected_data(),
    y_title = "Age Standardized Incidence Rate (per 1000)",
    input$inc_disease, input$inc_region
  )
})

renderText({
  paste("The selected model was:", inc_selected_data()$model[1])
})
```
Ribbons: Blue = 95% confidence interval, 
Grey = 95% credible interval if RW1 or RW2 selected, 
otherwise 95% confidence interval if loess selected

## Age-Standardized Life Prevalence Rate

```{r Life Prev Rate Plot}
# Input -----------------------------------------------------------------------
inputPanel(
  selectInput("life_region",
    label = "CHSA",
    choices = sort(unique(orig_life_prev_df$HEALTH_BOUNDARIES)),
    selected = "1110 Fernie"
  ),
  selectInput("life_disease",
    label = "Disease",
    choices = unique(orig_life_prev_df$DISEASE),
    selected = "AMI"
  )
)

# Reactive Elements -----------------------------------------------------------
life_selected_data <- reactive({
  life_prev_df |>
    dplyr::filter(HEALTH_BOUNDARIES == input$life_region & DISEASE == input$life_disease)
})

orig_life_selected_data <- reactive({
  orig_life_prev_df |>
    dplyr::filter(HEALTH_BOUNDARIES == input$life_region & DISEASE == input$life_disease)
})

# Output -----------------------------------------------------------------------
renderPlot({
  line_plot(life_selected_data(), orig_life_selected_data(),
    y_title = "Age Standardized Life Prevalence Rate (per 1000)",
    input$life_disease, input$life_region
  )
})

renderText({
  paste("The selected model was:", life_selected_data()$model[1])
})
```
Ribbons: Blue = 95% confidence interval, 
Grey = 95% credible interval if RW1 or RW2 selected, 
otherwise 95% confidence interval if loess selected