---
title: 'BC Chronic Diseases trends: A joinpoint regression analysis'
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "[Mahmoodur Rahman](https://sites.google.com/view/rmahmood/)"
output:
  html_document:
    df_print: paged
    code_folding: hide
bibliography: joinpoint_ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


```{r Load libraries}
library(here)
library(tidyverse)
library(segmented)
library(broom)
library(modelr)
library(purrr)

path <- here("data", "processed")
```

## Rationale

 Segmented or broken-line models are part of regression models where the
associations between the outcome and independent variable are
piece-wise linear, which means when two or more straight lines
connected at unknown values. These values are conventionally termed as
breakpoints, change-points or even join-points. Here the response
variable are different standardized chronic disease rates, and Time in
years (2001 to 2020) being the explanatory variable. Broken-line
relationships are common in many fields, including epidemiology,
occupational medicine, toxicology, and ecology, where sometimes it is of
interest to assess threshold value where the effect of the co-variate
(Time in years) changes [@ulm1991statistical; @betts2007thresholds]. In
other words, this model is used here to estimate abrupt changes in rates
in particular points in time, rather than smoothing it. The idea is to
aid epidemiologists in drawing inferences by estimating these
change-points.
 
## Data Pre-processing

### Data loading and wrangling

 - Data from all diseases by regions from over the years are combined to make single dataframe
 - We excluded diseases with missing and Zero values

```{r data load}
joinpoint_df <- read_csv(here(path, "joinpoint_df.csv"))
head(joinpoint_df)
```
 

## Exploratory data analysis

 - The following graph plots Age-standardized Incidence Rates of Chronic Diseases in Kimberley CHSA in from 2001 to 2020

```{r eda, fig.width=14, fig.height=12}

joinpoint_df |>
  filter(RATE == "INCIDENCE" & HEALTH_BOUNDARIES == "Kimberley") |>
  ggplot(aes(x=YEAR, y=STD_RATE_PER_1000)) +
  geom_line() +
  facet_wrap(~DISEASE) +
  labs(x = "Years", y = "Age standardized Incidence Rate (per 1000)") +
  ggtitle("Age-standardized Incidence Rates by Diseases in Kimberley (2001 to 2020")

```

#### Observations

 - "Changelings" can be seen, and not all lines are linear or smooth
 - Each "Piece" is considered a linear relation between Rates and Time (Year)

## Building the Model on sample data

### Arguments 

 - Model : Segmented regression / Joinpoint regression / Piecewise regression
 - Response (y): Age-Standardized Incidence Rate
 - Predictor : Year
 - `segmented.lm` is used from the `segmented()` package
 - We are performing a joinpoint regression on the data of ALZHEIMER_DEMENTIA of Fairview CHSA
 
```{r linear regression}

sample_df <- joinpoint_df |>
  filter(
      RATE == "INCIDENCE" &
      HEALTH_BOUNDARIES == "Fairview" &
      DISEASE == "ALZHEIMER DEMENTIA"
  ) |> 
  na_if(0) |> 
  drop_na()

lm_model <- lm(log(STD_RATE_PER_1000)~YEAR, 
               na.action= na.exclude,
               data = sample_df)
seg_obj<-segmented(lm_model,seg.Z=~YEAR)

```

### Model daignostics

#### Regression coefficients

```{r}
print(summary(seg_obj))
```

##### Observations

 - Breakpoint is in Year 2009

#### Average Annual Percent (AAPC) change

To summarize piecewise linear relationships in segmented regression models.
```{r aapc}
print(aapc(seg_obj, wrong.se = FALSE))

```
##### Observations

 - Positive AAPC refers to a significant trend

### Plot

```{r}
plot(seg_obj, 
     conf.level=.95, 
     is=TRUE, isV=FALSE,
     col=1, shade = TRUE, 
     res=TRUE, 
     res.col=1, 
     pch=19,
     main = "Predicted trend of Age-standardized Incidence Rates",
     xlab = "Time in Years",
     ylab = "log(Age-standardized Incidence Rate")

```

#### Observations

 - Epidemiological emphasis: Possible causes can be explored looking at the "Breakpoint" of Year 2009. 

## Implications

 - We developed an R shiny dashboard where a tab contains an interactive plot 
 showing breakpoints in trends of diseases per CHSA
 - The results from the model is also saved in a dataframe for future use
 
### References
 
 
 
 
 
 
 
 