---
title: 'DRAFT: BC Chronic Diseases trends by Sex (2001–2020): A joinpoint regression analysis'
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: 'Mahmoodur Rahman'
output:
  html_document:
    df_print: paged
    code_folding: hide
bibliography: joinpoint_ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```
```{r libraries}
library(dplyr)
library(readr)
```


# Highlevel tasks:

## Literature review / Write Rationale: (Will write rationale of this approach)
List of papers:
 - segmented: an R Package to Fit Regression Models with Broken-Line Relationships. (@muggeo2008segmented)
 - Estimating regression models with unknown break-points. (@muggeo2003estimating)
 - Global increasing incidence of young-onset colorectal cancer across 5 continents: a joinpoint regression analysis of 1,922,167 cases (@lui2019global)
 - Spatial clusters and temporal trends of malignant melanoma mortality in Ecuador. (@nunez2020spatial)
 - Worldwide suicide mortality trends by firearm (1990–2019): A joinpoint regression analysis (@ilic2022worldwide)
 
## Data Pre-processing:

### Loading the data

```{r Load Data}
df <- read_csv("/Users/mahmood/UBCMDS/591_capstone/master_df.csv") |> 
    mutate( RATE = as.factor(RATE)) |> 
    mutate( DISEASE = as.factor(DISEASE)) |> 
    mutate( SEX = as.factor(SEX)) |> 
    mutate( HEALTH_BOUNDARIES = as.factor(HEALTH_BOUNDARIES)) |> 
    mutate( GEOGRAPHY = as.factor(GEOGRAPHY))

```
 
 















 
# References
 
 
 
 
 
 
 
 