---
title: 'DRAFT: BC Chronic Diseases trends: A joinpoint regression analysis'
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: 'Mahmoodur Rahman'
output:
  html_document:
    df_print: paged
    code_folding: hide
bibliography: joinpoint_ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

# Calling the libraries:

```{r libraries}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(segmented)
library(broom)
library(tidyr)
library(modelr)
library(purrr)
library(fst)
```

# Background / Methodology / Literature review / Rationale: (Will write rationale of this approach)

List of papers:

 - segmented: an R Package to Fit Regression Models with Broken-Line Relationships. (@muggeo2008segmented)
 - Estimating regression models with unknown break-points. (@muggeo2003estimating)
 - Global increasing incidence of young-onset colorectal cancer across 5 continents: a joinpoint regression analysis of 1,922,167 cases (@lui2019global)
 - Spatial clusters and temporal trends of malignant melanoma mortality in Ecuador. (@nunez2020spatial)
 - Worldwide suicide mortality trends by firearm (1990–2019): A joinpoint regression analysis (@ilic2022worldwide)
 
# Data Pre-processing:

## Loading data :

 - Adding a column to represent if the rows have Null, Zero and Missing values

```{r Load Data}
CHSA_df <- read_csv("/Users/mahmood/UBCMDS/591_capstone/master_df.csv") %>%
  mutate( RATE = as.factor(RATE)) %>% 
  mutate( DISEASE = as.factor(DISEASE)) %>%
  mutate( HEALTH_BOUNDARIES = as.factor(HEALTH_BOUNDARIES)) %>%
  mutate(anyTRUE = !if_any(everything(), ~. %in% c(NA,"",0)))

```
## Data wrangling:

 - To exclude Null/NA values and Diseases with Zero values

```{r}

main_CHSA_df <- CHSA_df %>%
  dplyr::select(-c(STD_UCL_95, STD_LCL_95)) %>%
  mutate(anyTRUE = ifelse((
    RATE == 'INCIDENCE' & (
      (DISEASE == 'JUVENILE ARTHRITIS') |
        (
          HEALTH_BOUNDARIES == 'Downtown Victoria/Vic West' &
            DISEASE == 'EPILEPSY'
        ) |
        (
          HEALTH_BOUNDARIES == 'Snow Country' &
            DISEASE == 'HAEMORR STROKE'
        ) |
        (HEALTH_BOUNDARIES == 'Oak Bay' &
           DISEASE == 'EPILEPSY') |
        (
          HEALTH_BOUNDARIES == 'Vanderhoof Rural' &
            DISEASE == 'HAEMORR STROKE'
        ) |
        (HEALTH_BOUNDARIES == 'Panorama' &
           DISEASE == 'HOSP STROKE') |
        (HEALTH_BOUNDARIES == 'Grand Forks' &
           DISEASE == 'HOSP TIA') |
        (
          HEALTH_BOUNDARIES == 'Agassiz/Harrison' &
            DISEASE == 'HOSP TIA'
        ) |
        (
          HEALTH_BOUNDARIES == 'Burns Lake South' &
            DISEASE == 'HOSP TIA'
        ) |
        (
          HEALTH_BOUNDARIES == 'West Cariboo' &
            DISEASE == 'PARKINSONISM'
        ) |
        (
          HEALTH_BOUNDARIES == 'Agassiz/Harrison' &
            DISEASE == 'RHEUMATOID ARTHRITIS'
        ) |
        (
          HEALTH_BOUNDARIES == 'Burnaby Southwest' &
            DISEASE == 'RHEUMATOID ARTHRITIS'
        ) |
        (
          HEALTH_BOUNDARIES == 'Panorama' &
            DISEASE == 'OSTEOARTHRITIS'
        ) |
        (HEALTH_BOUNDARIES == 'Snow Country' &
           DISEASE == 'MS') |
        (
          HEALTH_BOUNDARIES == 'Snow Country' &
            DISEASE == 'PARKINSONISM'
        ) |
        (
          HEALTH_BOUNDARIES == 'Snow Country' &
            DISEASE == 'SCHIZOPHRENIA'
        ) |
        (
          HEALTH_BOUNDARIES == 'Burns Lake South' &
            DISEASE == 'PARKINSONISM'
        ) |
        (
          HEALTH_BOUNDARIES == 'Vanderhoof Rural' &
            DISEASE == 'PARKINSONISM'
        ) |
        (HEALTH_BOUNDARIES == "Hudson's Hope" &
           DISEASE == 'MS') |
        (
          HEALTH_BOUNDARIES == "Tumbler Ridge" &
            DISEASE == 'PARKINSONISM'
        ) |
        (
          HEALTH_BOUNDARIES == "Fort Nelson Population Centre" &
            DISEASE == 'MS'
        ) |
        (
          HEALTH_BOUNDARIES == "Fernie" &
            DISEASE == 'ALZHEIMER_DEMENTIA'
        )
    )
  ) |
    ((
      RATE == 'HSC' & (HEALTH_BOUNDARIES == 'Castlegar' &
                         DISEASE == 'HOSP TIA EPI') |
        (
          HEALTH_BOUNDARIES == 'Enderby' &
            DISEASE == 'SCHIZOPHRENIA EPI'
        ) |
        (
          HEALTH_BOUNDARIES == 'North Thompson' &
            DISEASE == 'HOSP STROKE EPI'
        ) |
        (
          HEALTH_BOUNDARIES == 'Brookswood/Murrayville' &
            DISEASE == 'HAEMORR STROKE EPI'
        ) |
        (
          HEALTH_BOUNDARIES == 'Snow Country' &
            DISEASE == 'HAEMORR STROKE EPI'
        ) |
        (
          HEALTH_BOUNDARIES == 'Vanderhoof Rural' &
            DISEASE == 'HAEMORR STROKE EPI'
        )
    )) |
    ((
      RATE == 'LIFE_PREV' & (
        HEALTH_BOUNDARIES == 'West Cariboo' &
          DISEASE == 'JUVENILE ARTHRITIS'
      ) |
        (
          HEALTH_BOUNDARIES == 'Fort Nelson Population Centre' &
            DISEASE == 'JUVENILE ARTHRITIS'
        )
    )), FALSE, anyTRUE))

```


# Exploratory data analysis

## Plotting of Rate against Time on sample dataframe: 

The following graph plots Age-standardized Incidence Rates of Chronic Diseases in Kimberley CHSA in from 2001 to 2020 (Will label the graph)

```{r eda, fig.width=14, fig.height=12}
# sample data

main_CHSA_df %>%
  filter(RATE == "INCIDENCE" & HEALTH_BOUNDARIES == "Kimberley") %>%
  ggplot(aes(x=YEAR, y=STD_RATE_PER_1000)) +
  geom_line() +
  facet_wrap(~DISEASE) +
  labs(x = "Years", y = "Age standardized Incidence Rate (per 1000)") +
  ggtitle("Age-standardized Incidence Rates by Diseases in Kimberley (2001 to 2020")

```

### Observations: 

 - "Breakpoints" can be seen. Not smooth
 - Each "Piece" is considered a linear relation between Rates and Time (Year)
 
## Summery sattistics of rates:

```{r rates summary}

summary(main_CHSA_df$STD_RATE_PER_1000)

```

### Observations:

 - None of the three rates, namely, HSC Prevalence, Age-Standardized Incidence Rate, and  Age-Standardized Life Prevalence have any negative value

# Building the Model on sample data:

## Specifics :

 - Model : Segmented regression / Joinpoint regression / Piecewise regression
 - Response (y): Age-Standardized Incidence Rate
 - Predictor : Year
 - `segmented.lm` is used
 - Filtering ALZHEIMER_DEMENTIA and 3213 Fairview CHSA
 
```{r linear regression}

sample_df <- main_CHSA_df %>%
  filter(RATE == "INCIDENCE"&
    HEALTH_BOUNDARIES == "Fairview" &
           DISEASE == "ALZHEIMER DEMENTIA")

lm_model <- lm(log(STD_RATE_PER_1000)~YEAR, data = sample_df)
seg_obj<-segmented(lm_model,seg.Z=~YEAR)

```


## Model daignostics:

### Regression coefficients:

```{r}
print(summary(seg_obj))
```
#### Observations: 

 - Breakpoint is in Year 2009

### Average Annual Percent (AAPC) change:

To summarize piecewise linear relationships in segmented regression models.
```{r aapc}
print(aapc(seg_obj, wrong.se = FALSE))

```
#### Observations:

 - Positive AAPC refers to an increasing trend

## Plotting:

```{r}
plot(seg_obj, 
     conf.level=.95, 
     is=TRUE, isV=FALSE,
     col=1, shade = TRUE, 
     res=TRUE, 
     res.col=1, 
     pch=19,
     main = "Predicted Age-standardized Incidence Rates of ALZHEIMER_DEMENTIA in Fairview CHSA (2001 to 2020)",
     xlab = "Time in Years",
     ylab = "log(Age-standardized Incidence Rate")

```

#### Observations:

 - Epidemiological emphasis: Possible causes can be explored looking at the "Breakpoint" of Year 2009. 


# Fitting the Model on data frame: 

### Model fitting:

```{r fitting incidence}

nested_df <- main_CHSA_df %>%
  filter(anyTRUE==TRUE)%>%
  group_by(RATE, HEALTH_BOUNDARIES, DISEASE) %>%
  arrange(RATE, HEALTH_BOUNDARIES, DISEASE, YEAR) %>% 
  nest() %>%
  mutate(fit = map(data, ~segmented(lm(log(STD_RATE_PER_1000)~ YEAR,
               na.action= na.exclude,
               data=.), seg.Z = ~YEAR)),
         results = map(fit, augment))%>%
  unnest(results)

```

### Display of results:

```{r }

view_nested_df <- nested_df %>%
  dplyr::select(c(RATE,  
                DISEASE,
                HEALTH_BOUNDARIES,
                YEAR,
                `log(STD_RATE_PER_1000)`,
                psi1.YEAR,
                .fitted))  
head(view_nested_df)

```

### Saving data: 

```{r }

temp_df <- view_nested_df %>%
  mutate(join_fitted = exp(.fitted)) %>%
  dplyr::select(c(RATE,
                  DISEASE,
                  HEALTH_BOUNDARIES,
                  YEAR,
                  join_fitted))

final_df <- main_CHSA_df %>%
  full_join(temp_df, by = c("RATE", 
                            "DISEASE", 
                            "HEALTH_BOUNDARIES", 
                            "YEAR")) %>%
  rename(join_obs = STD_RATE_PER_1000) %>%
  mutate(
    RATE = recode_factor(
      RATE,
      'HSC' = "Active Healthcare Contact (HSC) Prevalence",
      'INCIDENCE' = "Incidence Rate",
      'LIFE_PREV' = "Lifetime Prevalence"
    ),
  )%>%
  dplyr::select(c(RATE,
                DISEASE,
                HEALTH_BOUNDARIES,
                YEAR,
                join_obs,
                join_fitted))


write.csv(final_df,"/Users/mahmood/UBCMDS/591_capstone/joinpoint_df.csv", row.names = FALSE)

write_fst(final_df, "/Users/mahmood/UBCMDS/591_capstone/joinfast.fst")
```



```{r}
# 
# # Split dataframe by DISEASE
# incidence_df_list <- split(final_incidence_df, list(final_incidence_df$DISEASE))
# 
# # Write out separate CSV for each DISEASE
# for (DISEASE in names(incidence_df_list)) {
#     write.csv(incidence_df_list[[DISEASE]], paste0("IncidenceRate/",DISEASE, ".csv"), row.names = FALSE)
# }
```


# References
 
 
 
 
 
 
 
 