---
title: 'DRAFT: BC Chronic Diseases trends: A joinpoint regression analysis'
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: 'Mahmoodur Rahman'
output:
  html_document:
    df_print: paged
    code_folding: hide
bibliography: joinpoint_ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```



```{r libraries}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(segmented)
library(broom)
library(tidyr)
library(modelr)
library(purrr)
library(knitr)
```


# Rationale:

#### Diseases have trends ovber time, which often is not a straight line
#### There may be ups and down in the trend of a diseas over time
#### Beyween two points, a linear relation can be considered with time
#### Well establish methodlogy in oncology, chronic diseases, ecology, clinical longitudinal trials and most rencetly, Covid 19
#### Joinpoint regression / segmented regression / peicewise regression
 
```{r Load Data}

## Data Pre-processing:

### Loading data

CHSA_df <- read_csv("/Users/mahmood/UBCMDS/591_capstone/master_df.csv") %>%
  mutate( RATE = as.factor(RATE)) %>% 
  mutate( DISEASE = as.factor(DISEASE)) %>%
  mutate( HEALTH_BOUNDARIES = as.factor(HEALTH_BOUNDARIES)) %>% 
  drop_na()

CHSA_df[CHSA_df == "" | CHSA_df == 0] <- NA
row.has.na <- apply(CHSA_df, 1, function(x){any(is.na(x))})
sum(row.has.na)
CHSA_df <- CHSA_df[!row.has.na,]

```



```{r subset}
### Susetting data:

# Split the data frames according to Rates:

CHSA_data_list <- split(CHSA_df, f = CHSA_df$RATE) 

HSC_CHSA_df <- CHSA_data_list[[1]] %>% 
  as.data.frame()%>%
  dplyr::select(-c(RATE, STD_UCL_95, STD_LCL_95))
  
Incidence_CHSA_df <- CHSA_data_list[[2]] %>% 
  as.data.frame() %>% 
  dplyr::select(-c(RATE, STD_UCL_95, STD_LCL_95)) %>%
  filter(!(HEALTH_BOUNDARIES=='4111 Downtown Victoria/Vic West' & 
             DISEASE=='EPILEPSY')) %>% 
  filter(!(HEALTH_BOUNDARIES=='5110 Snow Country' & 
             DISEASE=='HAEMORR_STROKE')) %>% 
  filter(!(HEALTH_BOUNDARIES=='4114 Oak Bay' & 
             DISEASE=='EPILEPSY')) %>% 
    filter(!(HEALTH_BOUNDARIES=='5232 Vanderhoof Rural' & 
             DISEASE=='HAEMORR_STROKE')) %>% 
  filter(!(HEALTH_BOUNDARIES=='2334 Panorama' & 
             DISEASE=='HOSP_STROKE')) %>% 
  filter(!(HEALTH_BOUNDARIES=='1260 Grand Forks' & 
             DISEASE=='HOSP_TIA')) %>%  
  filter(!(HEALTH_BOUNDARIES=='2150 Agassiz/Harrison' & 
             DISEASE=='HOSP_TIA')) %>%  
  filter(!(HEALTH_BOUNDARIES=='5222 Burns Lake South' & 
             DISEASE=='HOSP_TIA')) %>% 
  filter(!(HEALTH_BOUNDARIES=='1461 West Cariboo' & 
             DISEASE=='PARKINSONISM'))%>% 
  filter(!(HEALTH_BOUNDARIES=='2150 Agassiz/Harrison' & 
             DISEASE=='RHEUMATOID_ARTHRITIS'))%>% 
  filter(!(HEALTH_BOUNDARIES=='2223 Burnaby Southwest' & 
             DISEASE=='RHEUMATOID_ARTHRITIS'))%>% 
  filter(!(HEALTH_BOUNDARIES=='2334 Panorama' & 
             DISEASE=='OSTEOARTHRITIS'))%>% 
  filter(!(DISEASE=='JUVENILE_ARTHRITIS'))%>% 
  filter(!(HEALTH_BOUNDARIES=='5110 Snow Country' & 
             DISEASE=='MS'))%>% 
  filter(!(HEALTH_BOUNDARIES=='5110 Snow Country' & 
             DISEASE=='PARKINSONISM'))%>% 
  filter(!(HEALTH_BOUNDARIES=='5110 Snow Country' & 
             DISEASE=='SCHIZOPHRENIA'))%>% 
  filter(!(HEALTH_BOUNDARIES=='5222 Burns Lake South' & 
             DISEASE=='PARKINSONISM'))%>%
  filter(!(HEALTH_BOUNDARIES=='5232 Vanderhoof Rural' & 
             DISEASE=='PARKINSONISM'))%>%
  filter(!(HEALTH_BOUNDARIES== "5322 Hudson's Hope" & 
             DISEASE=='MS'))%>%
  filter(!(HEALTH_BOUNDARIES== "5313 Tumbler Ridge" & 
             DISEASE=='PARKINSONISM'))%>%
  filter(!(HEALTH_BOUNDARIES== "5331 Fort Nelson Population Centre" & 
             DISEASE=='MS'))%>%
  filter(!(HEALTH_BOUNDARIES== "1110 Fernie" & 
             DISEASE=='ALZHEIMER_DEMENTIA'))



LifeP_CHSA_df <- CHSA_data_list[[3]] %>% 
  as.data.frame() %>% 
  dplyr::select(-c(RATE, STD_UCL_95, STD_LCL_95)) 

```


## We are looking at Age-standardized Incidence Rates of Chronic Diseases in Kimberley CHSA in from 2001 to 2020

```{r eda, fig.width=14, fig.height=12}
# sample data

Incidence_CHSA_df %>%
  filter(HEALTH_BOUNDARIES == "1130 Kimberley") %>%
  ggplot(aes(x=YEAR, y=STD_RATE_PER_1000)) +
  geom_line() +
  facet_wrap(~DISEASE) +
  labs(x = "Years", y = "Age standardized Incidence Rate (per 1000)") +
  ggtitle("Age-standardized Incidence Rates by Diseases in Kimberley (2001 to 2020")

```

#### "Breakpoints" can be seen. Not smooth
#### Each "Piece" is considered a linear relation between Rates and Time (Year)
 
## Summery sattistics of rates:

```{r rates summary}

summary(HSC_CHSA_df$STD_RATE_PER_1000)
summary(Incidence_CHSA_df$STD_RATE_PER_1000)
summary(LifeP_CHSA_df$STD_RATE_PER_1000)

```
### Observations: 

#### None of the three rates, namely, HSC Prevalence, Age-Standardized Incidence Rate, and  Age-Standardized Life Prevalence have any negative value

## Building the Model:

### Specifics :

#### Model : Segmented regression (Joinpoint regression)
#### Response (y): Age-Standardized Incidence Rate
#### Predictor : Year
#### `segmented.lm` is used
#### Filtering ALZHEIMER_DEMENTIA and 3213 Fairview CHSA
 
```{r linear regression}

sample_df <- Incidence_CHSA_df %>%
  filter(HEALTH_BOUNDARIES == "3213 Fairview" &
           DISEASE == "ALZHEIMER_DEMENTIA")

lm_model <- lm(log(STD_RATE_PER_1000)~YEAR, data = sample_df)
seg_obj<-segmented(lm_model,seg.Z=~YEAR)

```


### Model daignostics:

#### Regression coefficients:

```{r}
print(summary(seg_obj))
```
### Observations: 

#### Breakpoint is in Year 2009

#### average annual per cent (AAPC) change:

To summarize piecewise linear relationships in segmented
regression models.
```{r aapc}
print(aapc(seg_obj, wrong.se = FALSE))

```
###Observations:

#### Positive AAPC refers to an significant changes is trend over time.
#### It is the weighted average of the Annual Percentage Changes each year 
#### APC is the differences in slopes between two lines meeting in a chnagepoint.

#### Plotting:

```{r}
plot(seg_obj, 
     conf.level=.95, 
     is=TRUE, isV=FALSE,
     col=1, shade = TRUE, 
     res=TRUE, 
     res.col=1, 
     pch=19,
     main = "Predicted Age-standardized Incidence Rates of ALZHEIMER in Fairview CHSA (2001 to 2020)",
     xlab = "Time in Years",
     ylab = "log(Age-standardized Incidence Rate")

```

### Observations: (Will elaborate later)

#### Epidemiological emphasis: Possible causes can be explored looking at the "Breakpoint" of Year 2009. 

# References
 
 
 
 
 
 
 
 